---
title: "Employee Satisfaction Survey"
author: "Shana Nasrin"
date: "2024-08-27"
output: html_document
---

#### Install the packages
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
```
#### Load the CSV File
```{r}
employee_sat <- read.csv("C:/Users/shana/OneDrive/Desktop/Employee Attrition.csv")
```

#### Take a look at the dataset 
```{r}
head(employee_sat)
```
### View the data and obtain statistical summaries. Ensure data types are appropriate and there is no missing data.Determine the outcome and input variables.
#### Structure of the Data Frame
```{r}
str(employee_sat)
```

#### Check for any missing values
```{r}
missing_values <- sum(is.na(employee_sat))
print(paste("total missing values:", missing_values))
```

#### Statistical Summary
```{r}
summary(employee_sat)
```

#### Remove all the missing values
```{r}
employee_sat_clean <- na.omit(employee_sat)
```

#### Statistical summary of cleaned data
```{r}
summary(employee_sat_clean)
```
### Use the whisker boxplot to determine whether there is an outlier in the data.
#### Visualize Satisfaction level using box plot
```{r}
ggplot(data = employee_sat_clean, aes(y = satisfaction_level)) + 
  geom_boxplot(fill = "green", color = "darkblue") + 
  labs(title = "The box plot of satisfaction level", y = "satisfaction level")
```

### Use the appropriate diagram to demonstrate the categorical variable in the data.
#### Distribution of Departments
```{r}
ggplot(data = employee_sat_clean, aes(x = dept)) +
  geom_bar(fill = "darkblue", color = "black") + 
  labs(title = "The distribution of departments", x = "Department") + 
  geom_text(stat = 'count', aes(label = ..count..), 
            position = position_stack(vjust = 1.2))
```

#### Distribution of Salary
```{r}
ggplot(data = employee_sat_clean, aes(x = salary)) +
  geom_bar(fill = "orange", color = "black") + 
  labs(title = "The distribution of Salary", x = "salary") + 
  geom_text(stat = 'count', aes(label = ..count..), 
            position = position_stack(vjust = 0.5))
```

### Explain to your manager what actually leads to differences in income (salary).
#### Convert the salary into an ordered factor
```{r}
employee_sat_clean$salary <- factor(employee_sat_clean$salary, 
                                    levels = c("low", "medium", "high"), ordered = TRUE)
```

#### Analyze relationship between salary and other variables
```{r}
salary_analysis <- employee_sat_clean %>%
  group_by(salary) %>%
  summarise(
    avg_satisfaction = mean(satisfaction_level),
    avg_evaluation = mean(last_evaluation),
    avg_projects = mean(number_project),
    avg_hours = mean(average_montly_hours),
    avg_time_at_company = mean(time_spend_company),
    promotion_rate = mean(promotion_last_5years)
  )
print(salary_analysis)
```

#### Visualize relationship between salary and other variables
```{r}
p1 <- ggplot(data = employee_sat_clean, aes(x = salary, y = last_evaluation)) +
  geom_boxplot(fill = "blue", color = "black") + 
  labs(title = "Salary VS Evaluation Score", x = "salary", y = "Last Evaluation Score") 

p2 <- ggplot(data = employee_sat_clean, aes(x = salary, y = number_project)) +
  geom_boxplot(fill = "blue", color = "black") + 
  labs(title = "Salary VS Number of Projects", x = "salary", y = "Number of Projects")

p3 <- ggplot(data = employee_sat_clean, aes(x = salary, y = time_spend_company)) +
  geom_boxplot(fill = "blue", color = "black") + 
  labs(title = "Salary VS Years at Company", x = "salary", y = "Years at Company")

p4 <- ggplot(data = employee_sat_clean, aes(x = salary, y = average_montly_hours)) +
  geom_boxplot(fill = "blue", color = "black") + 
  labs(title = "Salary VS Average Monthly Hours", x = "salary", y = "Average Monthly Hours")
```

#### Install gridExtra to view the plots in a grid manner
```{r}
library(gridExtra)
```

#### Arrange the plots in a grid manner
```{r}
grid.arrange(p1, p2, p3, p4, ncol = 2)
```

#### Salary distribution across departments
```{r}
dept_salary <- employee_sat_clean %>%
  group_by(dept) %>%
  summarise(
    low = sum(salary == "low")/n(),
    medium = sum(salary == "medium")/n(),
    high = sum(salary == "high")/n()
  ) %>%
  arrange(desc(high))

print(dept_salary)
```

#### Visualize salary distribution across departments
```{r}
ggplot(data = employee_sat_clean, aes(x = dept, fill = salary)) +
  geom_bar(position = "fill") +
  labs(title = "Salary distribution across different departments", x = "Departments", y = "Proportion") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Determine the relationship among the variables.
#### Install the package corrplot for correlation analysis
```{r}
library(corrplot)
```

#### Select numeric variables for correlation analysis
```{r}
num_vars <- employee_sat_clean %>%
  select(satisfaction_level, last_evaluation, number_project, average_montly_hours, 
         time_spend_company,Work_accident, promotion_last_5years)
```

#### calculate correlation matrix
```{r}
corr_matrix <- cor(num_vars)
print(corr_matrix)
```

#### Create correlation plot
```{r}
corrplot(corr_matrix, method = "color", type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45, addCoef.col = "black", 
         number.cex = 0.7, tl.cex = 0.7)
```


### Plot a necessary graph that you think it will aid in understanding the data.
#### Calculate average satisfaction levels by salary and department
```{r}
avg_sat <- employee_sat_clean %>%
  group_by(salary, dept) %>%
  summarise(average_satisfaction = mean(satisfaction_level), count = n(), .groups = "drop") %>%
  arrange(dept, salary)
print(avg_sat)
```

#### Calculate overall statistics of satisfaction
```{r}
overall_sat_stat <- employee_sat_clean %>%
  summarise(
    avg_satisfaction = mean(satisfaction_level),
    median_satisfaction = median(satisfaction_level),
    sd_satisfaction = sd(satisfaction_level)
  )
print(overall_sat_stat)
```

#### Satisfaction levels by salary and department
```{r}
ggplot(data = employee_sat_clean, aes(x = salary, y = satisfaction_level, fill = dept)) + 
  geom_boxplot() +
  facet_wrap(~ dept, scales = "free_x") +
  labs(title = "Satisfaction levels by salary and department", x = "Salary", 
       y = "Satisfaction level", fill = "Department") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
```

### Identify and discuss anything else interesting that you see in the data.
#### Calculate the project engagement
```{r}
project_engagement <- employee_sat_clean %>%
  group_by(number_project) %>%
  summarise(
    avg_satisfaction = mean(satisfaction_level),
    avg_evaluation = mean(last_evaluation),
    avg_monthly_hours = mean(average_montly_hours),
    count = n(),
    .groups = "drop"
  )
print(project_engagement)
```

#### calculate work-life balance
```{r}
work_life_balance <- employee_sat_clean %>%
  group_by(salary) %>%
  summarise(
    avg_satisfaction = mean(satisfaction_level),
    avg_monthly_hours = mean(average_montly_hours),
    count = n(),
    .groups = "drop"
  )
print(work_life_balance)
```

#### Analyze promotion effects
```{r}
promotion <- employee_sat_clean %>%
  group_by(promotion_last_5years) %>%
  summarise(
    avg_satisfaction = mean(satisfaction_level),
    avg_evaluation = mean(last_evaluation),
    avg_monthly_hours = mean(average_montly_hours),
    avg_time_spend = mean(time_spend_company),
    count = n(),
    .groups = "drop"
  )
print(promotion)
```


### Report the conclusions of your analysis by writing a simple explanation that assumes no knowledge of statistics.

1. Most employees work 200 hours per month, no matter their salary is. Employees with higher salaries are much happier even though they work similar hours to others.

2. Employees working on 3 to 5 projects are much happier than working on 6 or more projects. As project workload increases, the employee experience potential burnout.

3. Employees who got promotion in the last 5 years are happier than others who did not get the promotion. This depicts that career growth is important for job satisfaction.

4. Working long hours make employees feel less satisfied than others.
